\propertitle

### 2019-12-08 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-07T21:28:11+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The best experimental measurements of instantaneous neuronal activity use
remarkable technologies, but can still only record a very small sample of
neurons -- hundreds at most -- compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only: they neglect that the recorded neurons are a sample from
a larger population. Such isolating assumptions sometimes escape attention,
being subtly hidden in the mathematics. Some probabilistic models try to
take unrecorded neurons into account, but they do so by describing each
neuron individually, thus becoming very complex
\citep{huang2015,battistinetal2017}. It would be useful to explore models
that operate in between, addressing the activity of the larger brain area
whence the sample comes, but \emph{collectively}, that is, without asking
about individual neuronal details. Such intermediate models would be useful
at least for preliminary investigations, to help us to decide which
hypotheses to consider or discard for more complex and costly models, or to
suggest new hypotheses.

@@

The model can be viewed as a straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
to build initial probability distributions has been used for different
kinds of investigations about the neuronal activity of various brain areas
and about other phenomena of importance to the neurosciences, for example
gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
However, our model can also be viewed as an approximation of a more
detailed model within a fully Bayesian approach; we shall constantly keep
this point of view in mind in our discussion.

@@

% Before applying the formula above to two concrete data sets we want to add
% two remarks about the maximum-entropy method that are seldom made in the
% neuroscientific literature. They are important for the interpretation of
% the results and are further discussed in \sect~\ref{sec:assumptions}.
% First, the maximum-entropy
% method % (based on the \emph{Shannon} relative
% % entropy, as opposed to other entropy functions such as Burg's
% % \citeyear{burg1975})
% rests on some implicit assumptions about the probabilities for the long-run
% frequency distribution of activities, besides the assumptions just
% mentioned at the end of the previous paragraph. So the often-heard
% statement that it gives \enquote{the maximally unbiased (or non-committal)
%   distribution} must be taken with a grain of salt.
% \citep{jaynes1986d_r1996,portamana2009,portamana2017}. Second,
A maximum-entropy distribution like $\px(\yA \| M, N)$ is equivalent to the
zeroth-order (Laplace
\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000})
approximation of four distinct distributions for the larger population,
which differ numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item\label{item:prob_freq} the most probable \emph{frequency} distribution
  for the total activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item\label{item:prob_binin} the \emph{belief} distribution for the value
  of the total activity at any time bin among the \emph{recorded} ones,
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item\label{item:prob_longrun} the most probable \emph{frequency}
  distribution for the total activity in a very long run of \emph{new} time
  bins, %$\ynuu(\yA)$,
\item\label{item:prob_binout} the \emph{belief} distribution for the value
  of the total activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
Their mathematical relationship is presented more explicitly in
appendix~\ref{sec:inference_prob}. The
distribution~\eqref{eq:app_maxent_pop_first} can therefore be interpreted
in each of the four ways above, in this approximate sense. In the present
case the validity of the approximation decreases as the ratio $nN/T$
increases. The most robust interpretation is~\ref{item:prob_freq}; and this
is how we prefer to interpret the distribution $\px$: as \emph{the most
  probable relative-frequency distribution} of the activity $\yA$. See
\sect~\ref{sec:assumptions} for further discussion about assumptions and
approximations.

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][-- we're only citing early papers using
these terms]{bohteetal2000,amarietal2003} of neuronal activity. Most, if
not all, such applications use the method \emph{at the sample level}. In
this section we discuss how our proposed application bears on this kind of
quantification and compare it with the traditional application at the
sample level. But such a comparison involves some methodological
caveats which we wish to discuss first.% For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

@@

\begin{figure}[!p]
\centering
\subcaption[]{\label{fig:plots_2_4mom_a}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_twomom_distributions10000_priorc.pdf}%
\\[2em]%
\subcaption[]{\label{fig:plots_2_4mom_b}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_twomom_distributions10000_prior2.pdf}%
\\[2em]%
\caption{Maximum-entropy distributions for the normalized total activity of
  the larger population, assuming two different sets of constraining moments
  and a larger population size $N=10\,000$.}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
These informational differences shows clearly in the shapes of the
distribution themselves, plotted in 
\fig~\ref{fig:plots_2_4mom_a}. The two-moment distribution
(\textcolor{myred}{dashed red}) is unimodal, the four-moment distribution
(\textcolor{myblue}{solid blue}) is bimodal. The five-moment distribution
would not be distinguishable from the four-moment one.  Figure~\ref{fig:plots_2_4mom_b} shows the corresponding distributions for the second data set;
also in this case they are visually very distinct.

@@

It is also possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic meaning~\eqref{eq:bayes_factor} to compare the probability of
the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of four-moment
sufficiency at the larger-population level, $N=10\,000$, versus the
hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart,
\end{equation}
that is, the data are 17 orders of magnitude more probable under the first
hypothesis than under the second. This result is also shown in
\fig~\ref{fig:comparisons_marginals}, where we see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).

@@

The second assumption, typical of maximum-entropy in general, concerns the
form of the probability distribution for the long-run frequencies ensuing
from the first assumption. It is taken to be either an \enquote{entropic
  prior} \citep{skillingetal1984,rodriguez1991,neumann2007}, which takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
moments under study \citep{portamana2017}. This assumption is crucial for
deriving the maximum-entropy based on with the Shannon entropy. A different
density leads to alternative maximum-entropy methods, for example based on
the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of the preceding sections were insensitive
to the reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

This means that the maximum-entropy approximation presented here can be
favourably used to get an idea of what a fully probabilistic calculation
could yield. -- With a far lower computational cost: for example, for
$N=1\,000$ the maximum-entropy distribution can be calculated in ten
minutes on a laptop, whereas the full probability calculation takes several
days on a high-performance computing cluster.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-08T20:13:23+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The best experimental measurements of instantaneous neuronal activity use
remarkable technologies, but can still only record a very small sample of
neurons -- hundreds at most -- compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only: they neglect that the recorded neurons are a sample from
a larger population. Such isolating assumptions sometimes escape attention,
being subtly hidden in the mathematics. Some probabilistic models try to
take unrecorded neurons into account, but they do so by describing each
neuron individually, thus becoming very complex
\citep{huang2015,battistinetal2017}. It would be useful to explore models
that operate in between, addressing the activity of the larger brain area
whence the sample comes, but \emph{collectively}, that is, without asking
about individual neuronal details. Such intermediate models would be useful
at least for preliminary investigations, to help us  decide which
hypotheses to consider or discard for more complex and costly models, or to
suggest new hypotheses.

@@

The model can be viewed as a straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
to build initial probability distributions has been used for different
kinds of investigations about the neuronal activity of various brain areas
and about other phenomena of importance to the neurosciences, for example
gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
However, our model can also be viewed as an approximation of a more
detailed model within Bayesian theory; we shall constantly keep this point
of view in mind in our discussion.

@@

% Before applying the formula above to two concrete data sets we want to add
% two remarks about the maximum-entropy method that are seldom made in the
% neuroscientific literature. They are important for the interpretation of
% the results and are further discussed in \sect~\ref{sec:assumptions}.
% First, the maximum-entropy
% method % (based on the \emph{Shannon} relative
% % entropy, as opposed to other entropy functions such as Burg's
% % \citeyear{burg1975})
% rests on some implicit assumptions about the probabilities for the long-run
% frequency distribution of activities, besides the assumptions just
% mentioned at the end of the previous paragraph. So the often-heard
% statement that it gives \enquote{the maximally unbiased (or non-committal)
%   distribution} must be taken with a grain of salt.
% \citep{jaynes1986d_r1996,portamana2009,portamana2017}. Second,
A maximum-entropy distribution like $\px(\yA \| M, N)$ is equivalent to the
zeroth-order (Laplace
\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000})
approximation of four distinct distributions for the larger population,
which differ numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item\label{item:prob_freq} the most probable \emph{frequency} distribution
  for the total activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item\label{item:prob_binin} the \emph{belief} distribution for the value
  of the total activity at any time bin among the \emph{recorded} ones,
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item\label{item:prob_longrun} the most probable \emph{frequency}
  distribution for the total activity in a very long run of \emph{new} time
  bins, %$\ynuu(\yA)$,
\item\label{item:prob_binout} the \emph{belief} distribution for the value
  of the total activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
Their mathematical relationship is presented more explicitly in
appendix~\ref{sec:inference_prob}. The
distribution~\eqref{eq:app_maxent_pop_first} can therefore be interpreted
in each of the four ways above, in this approximate sense. In the present
case the validity of the approximation decreases as the ratio $nN/T$
increases. The most robust interpretation is~\ref{item:prob_freq}, and that
is how we prefer to interpret the distribution $\px$: as \emph{the most
  probable relative-frequency distribution} of the activity $\yA$. See
\sect~\ref{sec:assumptions} for further discussion about assumptions and
approximations.

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][-- we're only citing early papers using
these terms]{bohteetal2000,amarietal2003} of neuronal activity. Most, if
not all, such applications use the method \emph{at the sample level}. In
this section we discuss how our proposed application bears on this kind of
quantification and compares with the traditional application at the sample
level. But such a comparison involves some methodological caveats which we
wish to discuss first.% For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

@@

\begin{figure}[!p]
\centering
\subcaption[]{\label{fig:plots_2_4mom_a}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_twomom_distributions10000_priorc.pdf}%
\\[2em]%
\subcaption[]{\label{fig:plots_2_4mom_b}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_twomom_distributions10000_prior2.pdf}%
\\[2em]%
\caption{Maximum-entropy distributions for the normalized total activity of
  the larger population, assuming two different sets of constraining moments
  and a larger population size $N=10\,000$.}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
These informational differences show clearly in the shapes of the
distribution themselves, plotted in 
\fig~\ref{fig:plots_2_4mom_a}. The two-moment distribution
(\textcolor{myred}{dashed red}) is unimodal, the four-moment distribution
(\textcolor{myblue}{solid blue}) is bimodal. The five-moment distribution
would not be distinguishable from the four-moment one.  Figure~\ref{fig:plots_2_4mom_b} shows the corresponding distributions for the second data set;
also in this case they are visually very distinct.

@@

It is also possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic interpretation~\eqref{eq:bayes_factor} to compare the probability of
the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of four-moment
sufficiency at the larger-population level, $N=10\,000$, versus the
hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart.
\end{equation}
% that is, the data are 17 orders of magnitude more probable under the first
% hypothesis than under the second. 
This result is also shown in
\fig~\ref{fig:comparisons_marginals}, where we see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).

@@

The second assumption, typical of maximum-entropy in general, concerns the
form of the probability distribution for the long-run frequencies ensuing
from the first assumption. It is taken to be either an \enquote{entropic
  prior} \citep{skillingetal1984,rodriguez1991,neumann2007}, which takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
moments under study \citep{portamana2017}. This assumption is crucial for
deriving the maximum-entropy based on the Shannon entropy. A different
density leads to alternative maximum-entropy methods, for example based on
the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}. The entropic prior depends on a reference
distribution and on a coefficient $L$ which determines the sharpness of the
prior's peak. The computations of the preceding sections were insensitive
to the reference distribution in their significant digits. We used three
biologically reasonable alternatives: a flat distribution, one linearly
decreasing with $\yA$, and a normal one with a broad peak at low
activities.

@@

This means that the maximum-entropy approximation presented here can be
favourably used to get an idea of what a fully probabilistic calculation
could yield, but with a far lower computational cost. For example, for
$N=1\,000$ the maximum-entropy distribution can be calculated in ten
minutes on a laptop, whereas the full probability calculation takes several
days on a high-performance computing cluster.
